---
title: 'Reproducible Research: Peer Assessment 1'
author: "Luis Urbina"
date: "Saturday, May 16, 2015"
output: html_document
---

## Loading and preprocessing the data
### 1.1)
Given the data file 'activity.csv', the data is loaded using 
the `read.csv()` function. The new dataset is named **da**.

```{r}
    wd <- getwd()
    if(!is.null(wd)) setwd(wd)
    da <- read.table(unz("activity.zip", "activity.csv"), header=T, sep=",")
```

### 1.2)
In order to know the dataset behavior, a short exploratory data analysis
of the dataset is done using the `str()`function.

```{r}
    str(da)
```

The data frame has 17,568 observaciones and three variables. The 
variables steps and interval are the 'integer' data type. The 
variable steps has also NA values. There is not need to transform 
or process the **da** data frame in order to perform the analysis.


## What is mean total number of steps taken per day?
### 2.1)
The NA values are going to be ignnored in this part of the analysis. 
To calculate the total number of steps taken per day, the `aggregate()` 
function is used. The new data frame is named **tidy.steps**.

The **total number of steps per day** is shown below. Notice that 
there are some days that do not have counted steps and they are not 
listed. In total are listed only 53 days, information obtained using 
the `dim()` function.

```{r,echo=TRUE}
    tidy.steps <- aggregate(.~date, data=da, sum)
    tidy.steps[,c(1,2)]
    dim(tidy.steps)
```

### 2.2) Histogram
A **histrogram of the total number of steps taken each day** is build 
and shown next. Notice that the maximum frequency (y-axes) is around 
10 according to the graphics. 

```{r,echo=TRUE}
    hist(tidy.steps$steps,breaks=20,col='coral',
         xlab='Steps Per Day',
         main='Histogram') 
```

### 2.3) Mean and Median 
Following is reported the **mean** of the total number of 
**steps taken per each day**. Results are shown:

```{r,echo=TRUE}
    tidy.date.mean <- aggregate(.~date, data=da, mean)
    tidy.date.mean[,c(1,2)]
```

It is also reported the **median** of the total number of 
**steps taken per each day**. Results are shown:

```{r,echo=TRUE}
    tidy.date.median <- aggregate(.~date, data=da, median)
    tidy.date.median[,c(1,2)]
```


## What is the average daily activity pattern?
### 3.1) Time Series Plot
It is calculated the **mean of the number of steps taken for each 
5-minute interval**. The new data frame is named **tidy.interval**. 
It is determined it has 288 observations using the `dim()` function.

```{r,echo=TRUE}
    tidy.interval <- aggregate(.~interval, data=da, mean)
    dim(tidy.interval)
```

A **time series plot** of the 5-minute interval (x-axis) and the 
average number of steps taken, averaged across all days (y-axis) 
is shown next:

```{r,echo=TRUE, fig.align='center'}    
    plot(tidy.interval[,c(1,2)], type='l', col='red',
         main='Time Series',
         xlab='5-Minute Interval (Day)',
         ylab='Steps (Average)')
```

### 3.2) Maximum Number of Steps
The **maximum number of steps** each 5-minute interval, on average across 
all the days, is calculated next.

```{r,echo=TRUE, results='asis'}    
    max(tidy.interval[,2])
```



## Imputing missing values
### 4.1) Total Number of Missing Values
To calculate the **total number of missing values** in the dataset, it is 
used the `apply()` function. First it is identified the rows with NA 
values, and next they are counted using the `sum()` funciton. 

```{r,echo=TRUE}    
    na.identifier <- apply(da, 1, function(x){any(is.na(x))})
    total.na <- sum(na.identifier)
    total.na
```


The total number of missing values is 2304.
 

### 4.2) Strategy for Imputing Missing Data
The strategy to fill the missing values is: 

a) first, calculate the mean for the 5-minute interval

b) second, replace the 5-minute interval mean in each missing value using a for-loop. It creates a new data set.

Firs, it is calculated the mean for the 5-minute interval. It is named
the **interval.mean**. Code is next:

```{r,echo=TRUE}    
    interval.mean <- mean(tidy.interval$steps)
    interval.mean
```

Now this interval.mean value will be replaced in any NA value in the 
dataset **da**, using a `for()` loop. 

### 4.3) Replacing Missing Values
A new dataset is created in which all the NA values in **da** have 
been replaced using a for-loop. The new dataset is named **da.full**. 
The algorithm is next:

```{r,echo=TRUE}
    da.full <- da
    for (i in 1:(length(da.full[,1]) )) {
        if (is.na(da.full$steps[i])) 
            da.full$steps[i] <- interval.mean
    }
```


### 4.4) What is the impact of imputing missing data on the estimates of the total daily number of steps?
### 4.4.1) Histogram After Replaced Missing Values
With replacement of missing values, the frequency of the steps per day
have increased, keeping the same distribution. As can be see in the
graphics below, the top frequency is over 15. 

```{r,echo=TRUE}
    tidy.steps.full <- aggregate(.~date, data=da.full, sum)
    hist(tidy.steps.full$steps,n=20,col='coral', 
         xlab='Steps (Day)',
         main='Histogram With Replacement of Missing Values') 
```


### 4.4.2) Mean
```{r,echo=TRUE}
    tidy.date.mean.full <- aggregate(.~date, data=da.full, mean)
    tidy.date.mean.full[,c(1,2)]
```

Both averages, with replacement and without replacement, of the steps
media for interval are the same. Numeric comparison is shown below.

```{r,echo=TRUE}
    mean(tidy.date.mean$steps)
    mean(tidy.date.mean.full$steps)
```

### 4.4.3) Median
```{r,echo=TRUE}
    tidy.date.median.full <- aggregate(.~date, data=da.full, median)
    tidy.date.median.full[,c(1,2)]
```

The averages of the median change, with replacement and without 
replacement, of the steps median for interval. Numeric comparison is 
shown below.


```{r,echo=TRUE}
    mean(tidy.date.median$steps)
    mean(tidy.date.median.full$steps)
```



## Are there differences in activity patterns between weekdays and weekends?
### 5.1) Algorithem

The next algorithm is going to create a new factor variable in the 
**da.full** dataset with two levels, "weekday" and "weekend", indicating
whether a given date is a weekday or weekend day. This new variable 
is named **day**. For this purpose, it is used the `weekdays()` function.

```{r,echo=TRUE}    
    da.full$aux <- weekdays(as.Date(da.full$date))
    days <- c('Monday','Tuesday','Wednesday','Thursday','Friday')
    da.full$day <- ""
    
    for (i in 1:(length(da.full[,1]) )) {
        if (da.full$aux[i] %in% days) 
            da.full$day[i] <- "weekday"
        else
            da.full$day[i] <- "weekend"
    }
```

Below is shown few rows of the new dataset **da.full** showing the new
created factor **day** and the dataset dimension.

```{r,echo=TRUE}
    dim(da.full)
    head(da.full)
```

### 5.2) Panel Plot to Compare Interval Acrross Weekdays and Weekends
A **panel plot** containing a time series plot fo the 5-minute interval and the average number of steps taken, average, is shown below.

```{r,echo=TRUE, fig.align='center'}
    require(stats)
    require(lattice)
    da.weekend <- aggregate(.~interval, 
                     data=da.full[da.full$day=='weekend',c(1,2,3)],mean)
    da.weekday <- aggregate(.~interval, 
                     data=da.full[da.full$day=='weekday',c(1,2,3)],mean)
    
    da.weekend$day <- 'weekend'
    da.weekday$day <- 'weekday'
    
    da.day <- rbind(da.weekday, da.weekend)
    xyplot(steps~interval|day, data=da.day, type='l', col='blue',
           layout=c(1,2), main='Time Series',
           xlab='Interval',ylab='Number of Steps (Avg.)')
```